Probabilistic Inference of NYC Congestion from Taxi Data

### 2017-12-10 reports/paper/paper.tex
related work

\usepackage{times}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algorithmic,inconsolata}
\usepackage{lipsum}
\usepackage{todonotes,tikz}
\usepackage{cite}
\usepackage{natbib}
\usepackage{threeparttable}
\newcommand{\sd}{\operatorname{SD}}

@@

@@

\begin{algorithm}
  \begin{algorithmic}
    \STATE{\COMMENT{Optimizing using \texttt{torch.optim.Adam}}}

@@

    Update $\bm W$.
    } \ENDFOR
  \end{algorithmic}
  \caption{Training algorithm for softmax path selection}
  \label{alg:main_alg}
\end{algorithm}

\subsection{Prediction}
\label{sub:predict}
We conclude this section by briefly discussing prediction in our models. In the uniform model, prediction is straightforward. We calculate the expected time $\hat T = \sum_i w_i \pi_i$, where $\pi_i$ is the marginal probability that node $i$ is included in a uniformly randomly chosen path. Computing prediction is slightly more subtle in the softmax regression model. In the softmax regression model, we sample $L_{\text{pred}}$ uniformly chosen paths and take the weighted average of the sums of $w_i$ in each path, where the weights are proportional to $\exp(-\sum_{i \in Z} w_i)$ for path $Z$.

% {\color{gray}
% \begin{itemize}
% \item How do you plan on training your parameters / inferring the
%   states of your latent variables (MLE / MAP / Backprop / VI / EM / BP / ...)

% \item What are the assumptions implicit in this technique? Is it an approximation or exact? If it is an approximation what bound does it optimize?

% \item What is the explicit method / algorithm that you derive for learning these parameters?
% \end{itemize}

% }

@@

\subsection{Data Processing}

For our dataset, we started with the TLC dataset of all NYC taxi trips taken in
January 2009. From the raw data, we added several computed columns.
We one-hot-encoded columns for hour of day (24 columns) as well as for
day of the week (7 columns). We also explicitly computed the duration of the trip  in seconds as the dependent variable to predict.

In addition to adding columns, we filtered the original dataset. We only consider trips
whose start- and end-locations that falls inside a rectangle that bounds Manhattan. We also removed trips with duration less than two minutes or more than two hours.

We performed a principal component analysis on all de-meaned and
standardized geographic coordinates (both start
and end locations) from the filtered data. We thus transform the coordinates into the first and second principal components. This allowed us to compute the rotational matrix so that Manhattan lies approximately orthogonal to the $x$-axis. For the discrete
grid models, these coordinates were further discretized into a $20 \times 70$
grid. Each node in the grid approximately corresponds to a $400\text{ft} \times 400\text{ft}$ area. We chose the grid size as a compromise between computational tractability and model expressiveness.

We saved the dataset as a whole, and we also partitioned into $(\text{day of week},
\text{hour})$ groups, using the one-hot encoded vectors. Some models were
trained on the entire dataset (e.g.\ Neural Network), and some were trained
individually on the partioned datasets (e.g.\ Structural Models).

\subsection{Baselines}

We calculated three baselines. The first was a linear regression predicting
trip duration from trip distance (calculated using the Pythagorean Theorem and
the start and end coordinates), and trained on a partitioned data set.

The second was a neural network trained on the
entire dataset taking as input the start and end coordinates as well as the day
of week and hour one-hot encoded vectors. The neural network had one hidden
layer with 50 neurons and a $\tanh$ activation function.

The third baseline was constructed by querying the Google Maps Distance Matrix
API.\footnote{\url{https://developers.google.com/maps/documentation/distance-matrix/}}
While there was no need to train this baseline, there was a rate limit
constraint on how many test data points it could be tested on. We assess its
performance on 2,500 data points in a representative partioned test dataset.

\subsection{Training}

Training implementations varied from model to model. In training the neural
network baseline, SGD was used with a learning rate of $10^{-7}$ run for 12
epochs on the full data set. This took between 30 minutes and an hour. The
uniform paths structural model was similarly trained with SGD, with a learning
rate of $10^{-4}$.

The uniform model was only trained on a subset of the
dataset corresponding to all trips on a given day of the week and on a given
hour. Convergence appeared after roughly two epochs, and the model was run for 5
epochs taking just under ten minutes.

The softmax regression model was trained
using the Adam optimizer \cite{kingma2014adam} with learning rate $10^{-2}$ for 10 epochs,
taking approximately an hour. Additionally, the softmax regression model take hyperparameters $L$ (sample size in each stochastic draw) and $\sigma^2$ (variance of trip time). We chose $L = 20$ and $\sigma^2 = 100$. These values are again chosen to strike a bargain between realism and computation feasibility. 
% TODO: jchen confirm this for me:

@@

{\color{gray}\begin{itemize}
\item What were the results comparing previous work / baseline systems / your systems on the main task?
\item What were the secondary results comparing the variants of your system?
\item This section should be fact based and relatively dry. What happened, what was significant?
\end{itemize}
}

We train our model and test the model on an out-of-sample test-set using the methods discussed in Section~\ref{sub:predict}.

Table~\ref{tab:mainres} contains the main prediction results of our model relative to some benchmark predictions
such as the Google Maps Distance Matrix API, neural network, and linear regression.

\begin{table*}[h!]
\centering
\begin{threeparttable}
\begin{tabular}{lrrrrr}
\toprule
& \multicolumn{3}{c}{Baseline} & \multicolumn{2}{c}{Model} \\\cmidrule(r){2-4} \cmidrule(l){5-6} 
{} & Google Maps & Neural Network\tnote{2} & Linear Regression\tnote{3} & Uniform Model & Softmax Model \\
\midrule
$\sd(T)$ & 7.40 & 6.85& 6.13&  6.22 & 6.22 \\ [1em]
$\hat \E[\epsilon]\tnote{1}$ & -0.83 &-0.07 & -0.05&  0.61 & 0.99 \\
$\sd(\epsilon)$ & 4.98 & 3.88& 5.01&  4.28 & 4.28 \\[1em]
$\hat \E[|\epsilon|]$ & 3.42 & 2.44& 3.60&  2.82 & 2.86 \\
$\text{Median}(|\epsilon|)$ & 2.55 & 1.67& 2.93 &  1.92 & 1.92 \\
$\text{Percentile}(|\epsilon|, 99)$ & 16.54 &12.79 & 16.62&  14.72 & 15.24 \\ [1em]
Test-set\tnote{4}  \, $R^2$ & 0.54 & 0.68& 0.33 & 0.53 & 0.53 \\
\bottomrule
\end{tabular}
\caption{Prediction results on test set}
\label{tab:mainres}
\begin{tablenotes}
% \item[\textbf{Notes:}]
\item[1] Here $\epsilon = (\text{Actual trip duration}) - (\text{Predicted duration})$ in minutes. $\hat \E$ denotes the sample mean. 
\item[2] We restrict the test-set to 8:00--9:00 AM on Tuesdays in January 2009 for all models \emph{except} the neural network, for which we use an out-of-sample test set on the \emph{entire} range of trip times, since the neural network encodes time-of-day and day-of-week data. 
\item[3]In linear regression, we simply regress $T$ on straight-line distance between the source and the destination.  
\item[4] Test-set $R^2$ is calculated as \smash{$1-\frac{\var(\epsilon)}{\var(T)}$} on the underlying dataset being tested on. Note that the test-set $R^2$ is an out-of-sample $R^2$. 
\end{tablenotes}
\end{threeparttable}
\end{table*}

>>>>

\usepackage{times}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{lipsum}
\usepackage{todonotes,tikz}
\usepackage{cite}
\usepackage{natbib}

@@

Travel time prediction has historically been a topic of research interest. On the topic of freeway travel time prediction, \cite{van2002freeway} proposed a Recurrent Neural Network (RNN) approach in order to address the temporal aspect of traffic prediction; \cite{wu2004travel} adopted the Supportive Vector Regression (SVR) method applied to time-seires analysis and reduced prediction errors in cases where previous methods generate especially large errors. We notice that earlier papers share the common feature of using freeway data for training and prediction, which probably arised from the limited availability of GPS data sources. Nevertheless, we realize that even for freeways, which in general have less traffic conditions than city roads, there is evidence that there exists a high level of non-linearity in the data, which contributes to the good performance of models such as RNN and SVR.

With the ubiquity of taxi GPS data, recent work focused on estimating the traffic conditions in cities using sparce probe data and used travel time predictions as the metric as the performance of the model. \cite{hunter2009path} uses a Bayesian framework and used an expectation maximization algorithm that simultaneously learns the likely paths taken by probe vehicles as well as the travel time distributions through the network; \cite{herring2010estimating} used a Coupled Hidden Markov Model (CHMM) and determines the most likely path by allocating the travel time between two consecutive location observations to roads in the `M' step.

One thing that distinguishes our attempt to the previous work is that we have much more missing information. Instead of sparse probe data, we are only working with start and end locations, which presents much more challenges in actual path inference and parameter optimization. \cite{zhan2013urban} provided some very helpful insights since it used the same TLC dataset of taxi ride. It contructed a faithful representation of the manhattan streets network, but limited the discussion to a much smaller district in Manhattan to reduce computational complexity.

@@

{\color{gray}
\begin{itemize}
\item How do you plan on training your parameters / inferring the
  states of your latent variables (MLE / MAP / Backprop / VI / EM / BP / ...)

\item What are the assumptions implicit in this technique? Is it an approximation or exact? If it is an approximation what bound does it optimize?

\item What is the explicit method / algorithm that you derive for learning these parameters?
\end{itemize}

\begin{algorithm}
  \begin{algorithmic}
    \STATE{\COMMENT{Optimizing using Adam in PyTorch}}

@@

    Update $\bm W$.
    } \ENDFOR
  \end{algorithmic}
  \caption{Training algorithm for softmax path selection}
  \label{alg:main_alg}
\end{algorithm}
}

@@

For our dataset, we started with the TLC dataset of all NYC taxi trips taken in
January 2009. From the raw data, several computed columns were added.
One-hot-encoded columns were added for hour of day (24 columns) as well as for
day of the week (7 columns). The duration of the trip was also explicitly
computed in seconds as the dependent variable to predict.

In addition to adding columns, the original dataset was filtered. Geographic
coordinates were\ldots

@@

\begin{itemize}
\item What were the results comparing previous work / baseline systems / your systems on the main task?
\item What were the secondary results comparing the variants of your system?
\item This section should be fact based and relatively dry. What happened, what was significant?
\end{itemize}

\begin{table*}
  \centering
  \missingfigure{}
  \caption{This is usually a table. Tables with numbers are generally easier to read than graphs, so prefer when possible.}
  \label{fig:mainres}
\end{table*}
